### 2016-01-18 paper.tex
Refine introduction a bit

Sandia National Laboratories, in collaboration with the US Department of Energy
developed the Reference Model devices to provide modeling targets for advancing
marine hydrokinetic (MHK) energy technology \cite{Neary2014}. The Reference
Model 2 (RM2), a vertical-axis cross-flow turbine intended for river
environments, was initially designed by Barone et al. \cite{Barone2011} using
the CACTUS vortex line code \cite{Murray2011}. A scaled physical model of the
RM2 was then designed and built by the Saint Anthony Falls Laboratory (SAFL) to
validate the CACTUS predictions. Experimental performance measurements were
significantly lower than the numerical results \cite{Hill2014}, which was
assumed to be a result of the physical model's low Reynolds number
\begin{equation}
    Re_D = \frac{U_\infty D}{\nu},
\label{eq:Re}
\end{equation}
where $U_\infty$ is the free stream velocity, $D$ is the rotor diameter (note
that in some cases the blade chord length $c$ is used as this length scale), and
$\nu$ is the fluid kinematic viscosity.

As a consequence of the inherently unsteady nature of the cross-flow rotor, the
RM2 blades experience large variations in angles of attack as they rotate about
their axis (``cross-flow'' means the axis of rotation is perpendicular to flow
direction). These variations become larger at lower tip speed ratio
\begin{equation}
    \lambda=\frac{\omega R}{U_\infty},
\end{equation}
where $\omega$ and $R$ are the rotor angular velocity and radius
\cite{Para2002}. Typically, the maximum angles of attack are sufficiently large
that the blades operate under dynamic stall, which is a complex unsteady process
and deviates significantly from static foil behavior, during part of the
rotation. Furthermore, the higher the solidity of a cross-flow turbine, the
lower the optimal tip-speed ratio at which it operates. Marine Hydrokinetic
(MHK) cross-flow turbines typically have higher solidity than cross-flow wind
turbines (VAWT), and hence operate at lower tip-speed ratios. Since MHK turbines
operate in a higher density fluid, unsteady dynamic effects related to the
blades' pitching motion and flow curvature also become more important when
compared to wind turbines.

>>>>

The Reference Model Project (RMP), sponsored by the US Department of Energy,
produced six marine hydrokinetic (MHK) technology point designs as reference
models (RM) to serve as non-proprietary test articles for open research
development, and to benchmark performance and costs for technology developers
\cite{Neary2014, Neary2014a}. Open-source RMP products, along with supporting
documentation, are available at the RMP website
\url{http://energy.sandia.gov/rmp} to facilitate their use in future R\&D
studies by industry, academia, and national laboratories. These products
include: Technical specifications and computer-aided design (CAD) files for each
RM device to allow exact replication for physical and numerical modeling
studies; Resource site information used to design each RM device; and references
to physical modeling data sets that can be used to validate numerical modeling
design and analysis tools.

Reference Model 2 (RM2) is a dual-rotor, vertical-axis cross-flow river turbine
that was designed to operate in a reach of the lower Mississippi River near
Baton Rouge, Louisiana \cite{Barone2011, Neary2011}. The rotor has three blades,
a relatively low solidity $Nc/(\pi D)$ or chord-to-radius ratio $c/R$, and a
preliminary analysis with Sandia's CACTUS vortex line numerical model
\cite{Murray2011} predicted a full-scale performance coefficient of 47\% at 1
m/s flow speed and a tip speed ratio $\lambda \equiv \frac{\omega R}{U_\infty} =
3.15$ \cite{Barone2011}, where $\omega$ and $R$ are the rotor's angular velocity
and radius, respectively. An initial experimental measurement campaign with a
1:15 scale RM2 rotor conducted at the Saint Anthony Falls Laboratory (SAFL) at
the University of Minnesota resulted in maximum power coefficient of
approximately 5\% at a tip speed ratio $\lambda = 2.2$. This discrepancy between
numerical and physical model performance prompted the experimental measurements
presented here, namely due to the low Reynolds number---$Re_D = U_\infty D /
\nu$, where $U_\infty$ is the free stream velocity, $D$ is the rotor diameter,
and $\nu$ is the fluid kinematic viscosity---of the 1:15 scale tests, which were
performed at $Re_D \sim 10^5$ \cite{Hill2014}.

As a consequence of the inherently unsteady nature of the cross-flow rotor, the
RM2 blades experience large variations in angles of attack as they rotate about
their axis (``cross-flow'' means the axis of rotation is perpendicular to flow
direction). These variations become larger at lower tip speed ratio
\cite{Para2002}. Typically, the maximum angles of attack are sufficiently large
that the blades operate under dynamic stall, which is a complex unsteady process
and deviates significantly from static foil behavior, during part of the
rotation. Furthermore, the higher the solidity of a cross-flow turbine, the
lower the optimal tip-speed ratio at which it operates. Marine Hydrokinetic
(MHK) cross-flow turbines typically have higher solidity than their similar wind
turbine counterparts (often called vertical-axis wind turbines, or VAWTs), and
hence operate at lower tip-speed ratios.
